Monster or Lambda?  A likelihood analysis

### 2023-01-26 inst/monster_vs_lambda.Rmd
reorganisation: start simple

First we need to define a slight generalization of `home_away3()`:

@@

We will try on a simple case:

```{r,label=try2x2}
M <- matrix(c(
    NA, 1+5i, 3+11i, NA),
    nrow=2,ncol=2,byrow=TRUE)

teams <- LETTERS[seq_len(2)]
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
M
home_away(M)
ha3(M,lambda=1.88)
```

(maybe one day I will include this function in the `hyper2` package).
Copying from `home_advantage.Rmd`:

@@

```{r,syntheticdatapp}
set.seed(0)
P <- matrix(0,4,4)
teams <- LETTERS[1:4]
dimnames(P) <- list("@home" = teams,"@away"=teams)
P
n <- 400 # number of matches between a pair 
M <- 0.15  # home monster
strengths <- 1/(1:4) # Zipf
strengths <- strengths/sum(strengths)
names(strengths) <- teams
strengths
for(home_team in 1:4){
  for(away_team in 1:4){
    if(home_team != away_team){
      prob_home_win <- (strengths[home_team] + M)/(strengths[home_team] + strengths[away_team] + M)
      home_wins <- rbinom(1,n,prob_home_win)
      home_losses <- n - home_wins
      P[home_team,away_team] <- home_wins + 1i*home_losses
    }
  } 
}
diag(P) <- NA
P
HP <- home_away(P)
```

@@

```{r,label=maxpmonsteq1,cache=TRUE}
mH <- maxp(HP,n=3,give=T)
mH
```

```{r proflikesyntha,cache=TRUE}
home <- seq(from=0.12,to=0.16,len=15)
supp_lambda_eq_0 <- profsupp(HP,"home",home,relative=FALSE)
```

```{r plotsynthprof}
plot(home,supp_lambda_eq_0,type='b')
plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
abline(h=c(0,-2))

```{r marginalsynthcalct,cache=TRUE}
f <- function(lambda){
 return(maxp(ha3(P,lambda=lambda),n=1,give=TRUE)$value)
}

lam <- seq(from=0.5,to=1.3,len=7)
supp_M_eq_0 <- sapply(lam,f)
```

@@

```{r plotbothtogether}
jj <- c(supp_M_eq_0,NA,supp_lambda_eq_0)
plot(jj-max(jj,na.rm=TRUE),type="b")
abline(h=c(0,-2))
```

>>>>

## Start simple: two teams

```{r,echo=FALSE}
jj <- matrix(c(NA, 1+2i ,4+5i, NA),2,2,byrow=TRUE)
teams <- LETTERS[1:2]
dimnames(jj) <- list("@home" = teams,"@away"=teams)
dimnames(jj) <- list("@home" = teams,"@away"=teams)
print(jj)
```

Above we see that A played at home 3 times, winning 1 and losing 2
[row 1 column 2 is $1+2i$]; B played at home 9 times, winning 4 and
losing 5 [row 2 column 1 is $4+5i$].  In its most general form, the
$2\times 2$ result matrix would be

$$
\begin{bmatrix}
-&\alpha+\beta i \\
\gamma + \delta i&-\\
\end{bmatrix}
$$

and the log-likelihood function would be, algebraically,

$$
\mathcal{S}=\log\mathcal{L}=
\log\left(\frac{A^\delta\cdot(A + M)^\alpha\cdot B^\beta
(B + M)^\gamma
}{
(A + B + M)^{\alpha+\beta+\gamma+\delta}
}\right)
=
\delta\log A + \alpha\log(1-B) + \beta\log B + \gamma\log(1-A)
$$

where $M$ is the strength of the home monster `home` and $A+B+M=1$, so
we understand that $M=1-A-B$.

$$
\frac{\partial\mathcal{S}}{\partial A}=\frac{\delta}{A}-\frac{\gamma}{1-A}\qquad
\frac{\partial\mathcal{S}}{\partial B}=\frac{\beta }{B}-\frac{\alpha}{1-A}
$$

At a maximum, therefore,

$$
\hat{A}=\frac{\delta}{\delta + \gamma}\qquad
\hat{B}=\frac{\beta }{\alpha  + \beta}
$$

Just to check, we evaluate the Hessian matrix and find

$$H=-
\left(\begin{array}{cc} 
\frac{\partial^2\mathcal{S}}{\partial A\partial A} & 
\frac{\partial^2\mathcal{S}}{\partial A\partial B} \\
\frac{\partial^2\mathcal{S}}{\partial B\partial A} & 
\frac{\partial^2\mathcal{S}}{\partial b\partial B} 
\end{array}\right)
=
\left(\begin{array}{cc}
\delta A^{-2} + \gamma(1-A)^{-2}&0\\
0&\beta B^{-2} + \alpha(1-B)^{-2}
\end{array}\right)
$$

which is clearly strictly positive-definite if $0<A,B<1$ and
$\alpha+\beta>0$, $\gamma+\delta>0$.

We will calculate numerical estimates using the
monster-parameterization for a simple case:

```{r,label=try2x2}
M <- matrix(c(
    NA, 5+1i, 9+2i, NA),
    nrow=2,ncol=2,byrow=TRUE)
teams <- LETTERS[seq_len(2)]
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
M
(Hsimp <- home_away(M))
maxp(Hsimp)
```

Compare the theoretical value of

```{r}
theoretical <- c(A=2/(2+9),B=1/(1+5))
fillup(theoretical,Hsimp)
```

showing good agreement.  We note that the model predicts a probability
of $\hat{A}/(\hat{A}+\hat{B})=\frac{12}{23}\simeq 0.52$ for $A$
winning in a match between $A$ and $B$ where the home team advantage
is not operating (on neutral ground, for example).

Now the $\lambda$ parameterization.  The likelihood function for the
same dataset is

$$
\mathcal{S}=\log\mathcal{L}=
\log\left(
\frac{(\lambda A)^\alpha B^\beta}{(\lambda A+B)^{\alpha+\beta}}
\cdot
\frac{A^\gamma (\lambda B)^\delta}{(A+\lambda B)^{\gamma+\delta}}
\right)$$

This is equal to $(\alpha+\delta)\log\lambda +(\alpha+\gamma)\log
A+(\beta+\delta)\log B -(\alpha+\beta)\log(\lambda A+B)
-(\gamma+\delta)\log(A+\lambda B)$ and even after simplifying [we may
assume $A+B=1$] the result does not appear to have a simple analytical
solution; we must proceed numerically.  To do so,

we need to define a
slight generalization of `home_away3()`:

@@

(maybe one day I will include function `ha3()` in the `hyper2`
package).

Now a more complicated example, a
modified version of the 4x4 matrix on `home_advantage.Rmd`:

@@

```{r,syntheticdatapp}
set.seed(0)
P <- matrix(0,4,4)
teams <- LETTERS[1:4]
dimnames(P) <- list("@home" = teams,"@away"=teams)
n <- 90 # number of matches between a pair 
M <- 0.15  # home monster
strengths <- (4:1)/10
strengths <- strengths/sum(strengths)
names(strengths) <- teams
strengths
for(home_team in 1:4){
  for(away_team in 1:4){
    if(home_team != away_team){
      prob_home_win <- (strengths[home_team] + M)/(strengths[home_team] + strengths[away_team] + M)
      home_wins <- rbinom(1,n,prob_home_win)
      home_losses <- n - home_wins
      P[home_team,away_team] <- home_wins + 1i*home_losses
    }
  } 
}
diag(P) <- NA
P
HP <- home_away(P)
```

@@

```{r,label=maxpmonstereq,cache=TRUE}
mH <- maxp(HP,n=3,give=T)
mH
```

```{r proflikesynthcc,cache=TRUE}
home <- seq(from=0.12,to=0.16,len=15)
supp_lambda_eq_1 <- profsupp(HP,"home",home,relative=FALSE)
```

```{r label=matchershower}
jjH <- ha3(P,lambda=1) # should match HP above
jjH
as.hyper3(HP) - jjH
```

```{r plotsynthprofa}
plot(home,supp_lambda_eq_1,type='b')
plot(home,supp_lambda_eq_1 - max(supp_lambda_eq_1),type='b')
abline(h=c(0,-2))
```

```{r showresult,cache=TRUE}
specificp.gt.test(ha3(P,lambda=1.4),"M",0)
specificp.gt.test(ha3(P,lambda=1.4),"M",1e-4)
```

```{r marginalsynthcale,cache=TRUE}
f <- function(lambda){
   o <- ha3(P,lambda=lambda)
   return(specificp.gt.test(o,"M",0)$null_support)
}

lam <- seq(from=1.9,to=2.4,len=5)
supp_M_eq_0 <- sapply(lam,f)
```

@@

```{r plotbothtogether}
jj <- c(supp_M_eq_0,NA,supp_lambda_eq_1)
plot(jj-max(jj,na.rm=TRUE),type="b")
abline(h=c(0,-2))
```
